---
title: "explore_pplot_data"
author: "Victoria Scholl"
date: "3/12/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
library(plyr)
library(dplyr)
library(sf)
library(ggplot2)
library(stringr) # for creating ribbon plot

# set working directory
setwd("~/github/jubilant-waffle/code/")

# load any local functions in external files 
source("supporting_functions.R")

# code for NEON site 
site_code <- 'NIWO'

# define the "bad bands" wavelength ranges in nanometers, where atmospheric 
# absorption creates unreliable reflectance values. 
bad_band_window_1 <- c(1340, 1445)
bad_band_window_2 <- c(1790, 1955)

#specify output directory 
out_dir <- paste0('../output/', site_code, '/')

# list of taxon IDs present at the NEON NIWO site
taxonList <- c("ABLAL","PICOL","PIEN","PIFL2")
```

# Exploring MRS-04 Data

## Permanent Plot measurements

```{r mrs4_pp_data}
# read the Excel file with MRS-04 Permanent Plot measurements
# I saved a .csv with the MRS4 sheet
mrs4_filename <- "../data/NIWO/tom_mrs04/Veblen_LTER_PP_Data_03262018_MRS04.csv"
mrs4_data <- read.csv(mrs4_filename)

# filter the data for entries with hc3 == 1
# hc3 column: Height class estimated during re-measurement in 2016.  
# 1 = main canopy, 2 = intermediate, 3 = subcanopy
mrs4_data_mainCanopy <- mrs4_data %>% dplyr::filter(hc3==1) %>% 
                          # convert from list to data frame 
                          as.data.frame() %>% 
                            # rename the Sp. and Tree. columns
                            dplyr::rename(taxonID = Sp.,
                                          TreeNum = Tree.)

# These taxon ID codes are different from the NEON taxon ID codes -
# these will be adjusted later. 

# count the number of trees per taxon ID in the main canopy
plyr::count(mrs4_data_mainCanopy, 'taxonID') %>%
  knitr::kable() %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover",
                                                  "condensed"),
                            full_width = F,
                            position = "left")
```

## Georeferenced tree point locations 

```{r}
# directory containing georeferenced shapefiles at MRS4
input_dir <- "../data/NIWO/tom_mrs04/mrs04_georeferenced_using_unavco_gps_points/"

# the "MRS4point.shp" feature class contains point locations
# for each numbered tree (TreeNum attribute)
mrs4point_layer <- "MRS4point.shp"
mrs4point <- sf::st_read(paste0(input_dir, mrs4point_layer))

# print the number of trees with stem locations in the MRS4point.shp file
print(paste0("There are ", as.character(length(mrs4point$TreeNum)), 
             " trees in the MRS4point shapefile"))

# plot all tree stem points
ggplot2::ggplot() + 
  ggplot2::geom_sf(data = mrs4point, size = 1, color = "black") + 
  ggtitle("All stem points within MRS4point.shp") + 
  coord_sf()

# The other shapefiles in the input_dir contain sapling measurements. 
# I will focus on the larger trees for now. 
```



```{r filter_treeNum_mrs4point}
# filter the tree point shapefiles for entries with matching tree numbers
# to the main canopy data (h3 == 1 from earlier)
mrs4point_mainCanopy <- mrs4point %>% 
              dplyr::filter(TreeNum %in% mrs4_data_mainCanopy$TreeNum)

# print the number of trees with stem locations and Main Canopy height class
print(paste0("There are ", as.character(length(mrs4point_mainCanopy$TreeNum)), 
             " trees in the MRS4point shapefile designated as the Main Canopy height class"))

ggplot2::ggplot() + 
  ggplot2::geom_sf(data = mrs4point_mainCanopy, size = 1, color = "black") + 
  ggtitle("MRS4point.shp filtered for Main Canopy Tree Numbers") + 
  coord_sf()

```

```{r newMRSpoint, eval=FALSE, include=FALSE}

# when the "newMRS4point.shp" file is used and checked,
# there are no trees that have the Main Canopy height class! 
# so I won't be using this shapefile for now. 

# the "newMRS4point.shp" feature class contains point locations
# for each numbered tree (TreeNum attribute)
mrs4point_layer <- "newMRS4point.shp"


mrs4point <- sf::st_read(paste0(input_dir, mrs4point_layer))

# print the number of trees with stem locations in the MRS4point.shp file
print(paste0("There are ", as.character(length(mrs4point$TreeNum)), 
             " trees in the MRS4point shapefile"))

# plot the tree points
ggplot2::ggplot() + 
  ggplot2::geom_sf(data = mrs4point, size = 1, color = "black") + 
  ggtitle("MRS4point.shp") + 
  coord_sf()

# filter the tree point shapefiles for entries with matching tree numbers
# to the main canopy data (h3 == 1 from earlier)
mrs4point_mainCanopy <- mrs4point %>% 
              dplyr::filter(TreeNum %in% mrs4_data_mainCanopy$TreeNum)

# print the number of trees with stem locations and Main Canopy height class
print(paste0("There are ", as.character(length(mrs4point_mainCanopy$TreeNum)), 
             " trees in the MRS4point shapefile designated as the Main Canopy height class"))

```


Now that we know which Tree Numbers are designated as the Main Canopy height class and also have stem locations in the shapefile, let's add a taxonID attribute to each point in the shapefile to keep track of its species. 

```{r taxonID_cleaning}
# create a look-up table (LUT) where the first column
# contains the taxonID codes and second column contains correspnding Tree Number.
taxonID_TreeNum_lut <- data.frame(taxonID = mrs4_data_mainCanopy$taxonID, 
                                  TreeNum = as.numeric(mrs4_data_mainCanopy$TreeNum))

# get the indices where the Tree Numbers in the point shapefile 
# match with Tree Numbers in the tree data 
idx <- match(mrs4point_mainCanopy$TreeNum, taxonID_TreeNum_lut$TreeNum)
mrs4point_mainCanopy$taxonID <- taxonID_TreeNum_lut[idx, "taxonID"]

# the species codes are different from the taxonID list used by NEON.
# let's adjust them to be the same. PIEN entries already match.  
# first, remove the factors so the taxonID values can be modified as characters.
mrs4point_mainCanopy$taxonID <- as.character(mrs4point_mainCanopy$taxonID)
mrs4point_mainCanopy$taxonID[mrs4point_mainCanopy$taxonID=="ABLA"] <- "ABLAL"
mrs4point_mainCanopy$taxonID[mrs4point_mainCanopy$taxonID=="PICO"] <- "PICOL"
mrs4point_mainCanopy$taxonID[mrs4point_mainCanopy$taxonID=="PIFL"] <- "PIFL2"

# filter the entries to species in the taxonID list 
# (remove the UNKN species entries)
mrs4point_mainCanopy <- mrs4point_mainCanopy %>% dplyr::filter(taxonID %in% taxonList)

# take another look at the number of trees per taxonID 
plyr::count(mrs4point_mainCanopy, 'taxonID') %>%
  knitr::kable() %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover",
                                                  "condensed"),
                            full_width = F,
                            position = "left")
```

Plot the stem points at MRS-04 designated as the Main Canopy height class and color them based on their species (taxon ID). 
```{r}
ggplot2::ggplot() + 
  ggplot2::geom_sf(data = mrs4point_mainCanopy, 
                   aes(colour=taxonID),
                   size = 1) + 
  ggtitle("MRS-04 Main Canopy stem points colored by species") + 
  coord_sf() +
  scale_color_brewer(palette = "Spectral")
```

```{r plot_stems_on_rgb}

# read RGB tile that contains the MRS-04 plot
rgb_filename <- "../data/NIWO/rgb/2017_NIWO_1_452000_4432000_image.tif"

rgb_red <- raster::raster(rgb_filename, band = 1) 
rgb_green <- raster::raster(rgb_filename, band = 2) 
rgb_blue <- raster::raster(rgb_filename, band = 3) 
rgb_stack <- raster::stack(rgb_red,rgb_green, rgb_blue)
rgb_cropped <- raster::crop(rgb_stack, as(mrs4point_mainCanopy, 'Spatial'))
raster::plotRGB(rgb_cropped, r = 1, g = 2, b = 3)
points(as(mrs4point_mainCanopy, 'Spatial'), pch = 16, cex = 0.5)
```


# Random Forest Classification

In theory, we know the species at each of the point locations in the plot above.
The code below extracts features from the remote sensing data cube and run the random forest classifier to predict species, then compare to the known species. 

```{r extract_features}
# read stacked AOP data
stacked_aop_filename <- "../data/NIWO/stacked_aop_data/stacked_aop_data_452000_4432000.rds"
stacked_aop_data <- readRDS(file = stacked_aop_filename)

# convert from SF to Spatial object to use the raster::extract function
mrs4point_mainCanopy_Sp <- sf::as_Spatial(mrs4point_mainCanopy,
                                  IDs = as.character(mrs4point_mainCanopy$TreeNum))

# TO DO: FILTER SHAPEFILE FOR ONLY THE TREES WITHIN THE CURRENT TILE

# extract all layers for each pixel in the shapefile
extracted_spectra <- raster::extract(stacked_aop_data, 
                                     mrs4point_mainCanopy_Sp, 
                                     df = TRUE)

extracted_spectra$TreeNum <- mrs4point_mainCanopy_Sp$TreeNum
extracted_spectra$taxonID <- mrs4point_mainCanopy_Sp$taxonID
extracted_spectra$X.1 <- 0
extracted_spectra$X <- 0
extracted_spectra$Y <- 0


# write extracted spectra to file 
extracted_features_filename <- paste0(out_dir, "MRS04_extracted_features.csv")
write.csv(extracted_spectra, 
              file = extracted_features_filename) 
```


```{r rf}
# read text file containing wavelengths 
wavelengths = as.numeric(unlist(read.table(paste0(out_dir,"wavelengths.txt"),
                                           sep="\n",
                                           skip = 1,
                                           col.names = 'wavelength')))

# read random forest classifier, variable name rf_model
load("../output/NIWO/rf_allSamplesPerClass_ntree5000_pca2InsteadOfWavelengths_nVar6/rf_model_allPolygons_halfDiameter.RData")

# read the values extracted from the data cube
df_orig <- read.csv(extracted_features_filename)
  
# Remove any spectra that have a height == 0
print(paste0(as.character(sum(df_orig$chm==0)), 
               " pixels have a height of 0 in the CHM"))
print("Removing these rows from the training set ... ")
```

